Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-04-05 pubs/tfp16/paper.tex
sync just paper
fix a few typos

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and non-strictness or laziness is more of a
nuisance--even a serious hindrance--than a convenience, much less ever
algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
of \emph{tying the knot}~\cite{tying-the-knot}.

@@

Much anecdotal evidence suggests that many others have similar difficulties
with semantics that are lazy (or non-strict) by default, with the most telling
recent concrete evidence being the push for, and realization of, the
\texttt{Strict} and \texttt{StrictData} language extensions newly available in
GHC 8.0.  Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practioners are by and large not computer
  scientists}, they are domain scientists (physicists, chemists, etc.) or
computational scientists specialized in numerical or numerically-oriented
algorithmic methods (e.g., Lagrangian, Eulerian, adaptive mesh refinement).
The concept of a space leak, for example, caused by other than a missing
explicit deallocation statement (C \emph{free}, etc.) is difficult to
understand and not something they particularly want to try to understand.
%Hence, we eschew explicit strictness annotations for varying uniform degrees
%of strictness.

@@

\item Haskell Block \\
  (ignored by MHS frontend, not other Haskell compilers;
  allows the same file to be tested on our compiler vs GHC)
\end{itemize}

@@

An almost unbounded amount of interesting work suggests itself; one such path
would lead to the recreation of GHC.  However, given the miniscule budget (and
thus the emphasis on student labor) we must maintain sharp focus on the main
goals:  implementing and evaluating mechanisms for automatic parallelism and
fault tolerance.

>>>>

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practitioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and non-strictness or laziness is more of a
nuisance--even a serious hindrance--than a convenience, much less ever
algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
of \emph{tying the knot}~\cite{tying-the-knot}.

@@

Much anecdotal evidence suggests that many others have similar difficulties
with semantics that are lazy (or non-strict) by default, with the most telling
recent concrete evidence being the push for, and realization of, the
\texttt{Strict} and \texttt{StrictData} language extensions newly available in
GHC 8.0.  Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practitioners are by and large not computer
  scientists}, they are domain scientists (physicists, chemists, etc.) or
computational scientists specialized in numerical or numerically-oriented
algorithmic methods (e.g., Lagrangian, Eulerian, adaptive mesh refinement).
The concept of a space leak, for example, caused by other than a missing
explicit deallocation statement (C \emph{free}, etc.) is difficult to
understand and not something they particularly want to try to understand.
%Hence, we eschew explicit strictness annotations for varying uniform degrees
%of strictness.

@@

\item Haskell Block \\
  (ignored by MHS front-end, not other Haskell compilers;
  allows the same file to be tested on our compiler vs GHC)
\end{itemize}

@@

An almost unbounded amount of interesting work suggests itself; one such path
would lead to the recreation of GHC.  However, given the minuscule budget (and
thus the emphasis on student labor) we must maintain sharp focus on the main
goals:  implementing and evaluating mechanisms for automatic parallelism and
fault tolerance.
